Prefix-free graphs and suffix array construction in sublinear space

### 2023-06-20 manuscript/main.tex
added some more figures
added some related work

\section{Prefix-free graphs}
% some intro?
The idea of prefix-free graphs is inspired by rsync \cite{}.
% what are prefix-free graphs
Given a set of sequences representing a pangenome, we can partition each
sequence into \emph{segment}s.
These segments form nodes of the prefix-free graph.
The edges of a graph represent the adjacencies of the segment pairs, which are 
present in the original strings.
The adjacent segments are overlapping by a fixed number of characters $k$.
We will call these $k$-sized overlaps triggers.
Another important feature of the segments is that they contain triggers only at
the start or at the end of a segment.
This feature guarantees that the set of segments is prefix-free, i.e. no
segment is a prefix of another segment.
% add sketch of a proof

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{images/pfg.png}
    \caption{Prefix-free graph in GFA format. Link lines omitted for brevity.}
    \label{fig:gfa}
\end{figure}

@@

\section{Related work}
MARIA \cite{2022maria}
Prefix-free Wheeler graphs \cite{2022pfwg}
Prefix-free parsing for building big BWTs \cite{2019boucher}
Rpair: rescaling repair with rsync \cite{2019gagie}

>>>>

\section{Prefix-free graphs}
% some intro?
The idea of prefix-free graphs is inspired by rsync \cite{}.
% what are prefix-free graphs
Given a set of sequences representing a pangenome, we can partition each
sequence into \emph{segment}s.
These segments form nodes of the prefix-free graph.
The edges of a graph represent the adjacencies of the segment pairs, which are 
present in the original strings.
The adjacent segments are overlapping by a fixed number of characters $k$.
We will call these $k$-sized overlaps triggers.
Another important feature of the segments is that they contain triggers only at
the start or at the end of a segment.
This feature guarantees that the set of segments is prefix-free, i.e. no
segment is a prefix of another segment.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{images/prefixfree_proof.png}
    \caption{Visual proof that the trigger words induce prefix-free segments.}
    \label{fig:proof}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{images/pfg_gfa.png}
    \caption{Prefix-free graph in GFA format. Link lines omitted for brevity.}
    \label{fig:gfa}
\end{figure}

% add graph figure

@@

\begin{figure*}
    \centering
    \includegraphics[width=\linewidth]{images/segment_table.png}
    \caption{}
    \label{fig:segment_sa}
\end{figure*}

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{images/segment_metainfo.png}
    \caption{}
    \label{fig:segment_meta}
\end{figure}

\section{Related work}
There are several proposed data structures implicitly relying on the structure
of prefix-free graphs.

Rpair: rescaling repair with rsync \cite{2019gagie}
Prefix-free parsing for building big BWTs \cite{2019boucher}
MONI and r-index \cite{}
Prefix-free Wheeler graphs \cite{2022pfwg}
MARIA \cite{2022maria}

All of these works implement their own version of prefix-free graphs.
We argue that separating prefix-free graphs as a standalone data structure can
bring several benefits:

\begin{itemize}
    \item reduce the complexity of the presentation
    \item allowing for better optimization of algorithms related to prefix-free graphs
    \item supporting the theoretical research by clearly deliminating the relevant terms
    \item allowing to bring prefix-free graphs closer to the biological data
\end{itemize}

Related to the last point, we remark that a possible future research can be in
the proper choice of trigger words.
As an example, we used stop codons in our experiments.
This could create prefix-free graphs capturing some biologically relevant phenomena.
Other options are recombination hotspots, highly repetitive elements, different
binding sites and sites with increased breakage.
